Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-06-24 survey/survey.tex
a bit on graphical lang

\documentclass{article}
\usepackage[utf8x]{inputenc}
\usepackage{agda}
\usepackage[nocenter]{qtree}
\usepackage{amssymb,amsthm,amsmath}
\usepackage{fullpage}
\usepackage{url}
\usepackage{comment}
\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{listings}
\usepackage{multicol}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usetikzlibrary{quotes}
\tikzset{>=latex}

@@

a graphical notation for programming in these models that reminiscent
of Penrose diagrams for categories~\cite{selinger-graphical}, Geometry
of Interaction (GoI) machines and Proof
Nets~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. In the graphical
notation, computation is modeled by the flow of particles in a
circuit.

\amr{
  Let’s have a geometric representation of spaces:
- A + B means putting A next to B
- A * B means putting A and B on two perpendicular axes and filling
the square (volume) between them
All the combinators would indeed have a natural action and if we have
a few nice pictures it would be great. What do you think? }

@@

%%%
\subsection{Spacetime Model}
\jc{While I really like this model, if we are going to keep it, we
really need a much beefier introduction. This is the first mention of
resources (and events and time). So we really should preface this by
saying that, once we have $\Pi$, there is another \emph{completely
different} but compatible interpretation in terms of resources. But
we should also think of whether this fits. I actually think that it
does fit in this paper, but perhaps not in this section. Maybe
also move to section 5?}

To gain an intuition for resources, we think of the types as possible
events, and of the values as consistent timelines of events. From this
perspective, $A * B$ denotes spatial composition, where both $A$ and
$B$ must happen in \emph{some} order, and $A + B$ denotes temporal
composition, where exactly one of $A$ and $B$ must happen at a given
time. The event $1$ is then trivial, and $0$ is impossible.  We
visualize the denotation of $1$ as a point in space, and $0$ as empty
space. The denotation of $A + B$ is the denotation of $A$ drawn on top
of the denotation of $B$ and separated by a border, and the denotation
of $A * B$ is just the denotation of $A$ written next to the
denotation of $B$. Now we can formally define a consistent timeline
recursively on the structure of a diagram:

\begin{itemize}
\item An endpoint (the denotation of $1$) has exactly one timeline
  through it.  Empty space (the denotation of $0$) has no timelines
  through it.
\item A timeline through a stack of alternatives
  $A_1 + A_2 + \cdots + A_n$ is a timeline through exactly one of the
  possible alternatives.
\item A timeline through a spatial composition of cells
  $A_1 * A_2 * \cdots * A_n$ is a timeline through \emph{all} of the
  cells, in some order.
\end{itemize}

For example, let $2$ be the type $1 + 1$. The following would be a
consistent timeline through the diagram for
$((2 + 2) * (2 + 2)) + ((2 + 2) * (2 + 2))$ times itself as shown in
the left two boxes. The right two boxes show the trade-off between
space and time encoded in distributivity. Since a timeline must pass
through a column in exactly one point, the two boxes on the right have
the same timelines:
\begin{center}
\includegraphics[scale=0.3]{images/tall-tl-1}
\quad
\includegraphics[scale=0.3]{images/tall-tl-2}
\qquad\qquad\qquad
\includegraphics[scale=0.3]{images/distrib-1}
\qquad
\includegraphics[scale=0.3]{images/distrib-2}
\end{center}

>>>>

\documentclass{article}
\usepackage[utf8x]{inputenc}
\usepackage{agda}
\usepackage[nocenter]{qtree}
\usepackage{amssymb,amsthm,amsmath}
\usepackage{fullpage}
\usepackage{url}
\usepackage{comment}
\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{listings}
\usepackage{multicol}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usetikzlibrary{quotes,fit,positioning}
\tikzset{>=latex}

@@

@@

% \jc{While I really like this model, if we are going to keep it, we
% really need a much beefier introduction. This is the first mention of
% resources (and events and time). So we really should preface this by
% saying that, once we have $\Pi$, there is another \emph{completely
% different} but compatible interpretation in terms of resources. But
% we should also think of whether this fits. I actually think that it
% does fit in this paper, but perhaps not in this section. Maybe
% also move to section 5?}

The textual representation of $\Pi$ is quite cumbersome: the
underlying principles however become quite intuitive when represented
in a more suggestive visual notation. There exist various graphical
notations for the monoidal categories that model $\Pi$ and related
languages including Penrose diagrams~\cite{selinger-graphical} and
Geometry of Interaction (GoI) machines and Proof
Nets~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. In this section, we
propose a novel graphical representation of $\Pi$ types and
combinators drawn from the resource usage embedded in the operational
semantics. As suggested by the operational semantics, the type $A * B$
can be viewed as denoting \emph{spatial composition}, where
\emph{both} $A$ and $B$ must happen in \emph{some} order or perhaps
simultaneously, and where $A + B$ denotes \emph{temporal composition},
where exactly \emph{one of} $A$ and $B$ must happen at a given time.

This idea leads to a resource-based interpretation of types as
\emph{possible events} and of the values as \emph{consistent timelines
  of events} as explained below. The event $1$ is trivial, and the
event $0$ is impossible.  We visualize the denotation of $1$ as a
point in space, and $0$ as empty space. The denotation of $A + B$ is
the denotation of~$A$ drawn on top of the denotation of $B$ and
separated by a border, and the denotation of $A * B$ is just the
denotation of $A$ written next to the denotation of $B$.  For example,
let $2$ be the type $1 + 1$ represented by a stack of two boxes each
with a point inside it. The type $(2 + 2) * (2 + 2)$ would then be two
such stacks laid next to each other. Composing further, the type
$((2 + 2) * (2 + 2)) + ((2 + 2) * (2 + 2))$ is built by stacking the
last two diagrams as shown below on the left:
\begin{center}
\hfill 
\includegraphics[scale=0.3]{images/tall-tl-1}
~
\includegraphics[scale=0.3]{images/tall-tl-2}
\hfill (*)
\end{center}
If we call the type on the left $A$ then the full diagram above
represents the type $A*A$.

To understand how diagrams (types) can transform through the
evaluation of  a program, we need to reason about the equivalence of
different diagrams. Consider for example, the two diagrams below:

\begin{center}
\hfill 
\includegraphics[scale=0.3]{images/distrib-1}
\qquad \qquad \qquad \qquad \qquad 
\includegraphics[scale=0.3]{images/distrib-2}
\hfill (**)
\end{center}

The diagram on the left represents the type $A * (B + C)$ while the
one on the right represents the type $(A * B) + (A * C)$. These two
types are equivalent in $\Pi$ but why are they equivalent in the
graphical notation? We argue that they admit the same consistent
timelines where a consistent timeline is defined recursively on the
structure of a diagram:
\begin{itemize}
\item An endpoint (the denotation of $1$) has exactly one timeline
  through it.  Empty space (the denotation of $0$) has no timelines
  through it.
\item A timeline through a stack of alternatives
  $A_1 + A_2 + \cdots + A_n$ is a timeline through exactly one of the
  possible alternatives.
\item A timeline through a spatial composition of cells
  $A_1 * A_2 * \cdots * A_n$ is a timeline through \emph{all} of the
  cells, in some order.
\end{itemize}

For example, the red line in the diagram (*) represents a consistent
timeline. For the diagrams (**), we note that any consistent timeline
for the left diagram must go through $A$ and either one of $B$ or $C$
and for the diagram on the right, the choice happens earlier but the
result is the same: any consistent timeline must go through either $A$
and $B$ or through $A$ and $C$.

% \amr{
%   Let’s have a geometric representation of spaces:
% - A + B means putting A next to B
% - A * B means putting A and B on two perpendicular axes and filling
% the square (volume) between them
% All the combinators would indeed have a natural action and if we have
% a few nice pictures it would be great. What do you think? }

% \newcommand{\graphzero}{}
% \newcommand{\graphone}{\ensuremath{\bullet}}
% \newcommand{\graphplus}[2]{\node (top) {#1}; \node [below of=top] {#2};}

% \begin{center}
% \begin{tikzpicture} %% [declare function = {plus(\x,\y) = \node (top){\x}; \node [below of=top] {\y};}]
% \node (x) {\graphone};
% \node (y) {\graphone};
% %% {plus(\x,\y)}
% \end{tikzpicture}
% \end{center}
